---
title: "Lab1"
output:
  html_document: default
  pdf_document: default
---

```{r setup}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(stats)
library(haven)
library(xtable)
library(knitr)

#Read in data from .dta file in current working directory:
df_2020 <- read_dta(file = "data_2020.dta")

```

Link to 2020 TimeSeries Dataset User Guide:
https://electionstudies.org/wp-content/uploads/2021/03/anes_timeseries_2020_userguidecodebook_20210324.pdf



----------------------------------------------------------------------
1. Are Democratic voters older or younger than Republican voters in 2020?

Respondent political affiliation: V201231x
Respondent Age: V201507x
```{r}
df <- df_2020[,c("V201231x","V201507x")] %>%
  mutate(
    dem_ind = ifelse(V201231x == 1 | V201231x == 2 | V201231x == 3, TRUE, FALSE)
  ) %>%
  filter(((V201231x < 4 & V201231x > 0) | (V201231x > 4 & V201231x < 8)) & (V201507x >= 18))
summary(df)
colnames(df) <- c('party_id', 'age', 'dem_ind')
party_age_hist <- df %>% 
  ggplot() + 
  aes(x = age, fill = dem_ind) + 
  xlim(18,max(df$age)) +
  geom_histogram(alpha = 0.5, position = 'identity', binwidth = 2) + 
  labs(
    title    = 'Political Party Affiliation Age Distributions',
    subtitle = 'Fill this in with the main takeaway of the plot.', 
    x        = 'Age',
    y        = 'Count of Voters', 
    fill     = 'Democrat'
  )

party_age_hist
t.test(df$age ~ df$dem_ind)
```

----------------------------------------------------------------------
2. Did Democratic voters or Republican voters report experiencing more difficulty 
voting in the 2020
election? (Exploratory, ungraded question: Were the types of reported difficulties 
the same, or different for respondents who identify with the different parties? 
We’re not going to grade this, and it isn’t a fully formed question here, but you [like us] 
might be interested in exploring this while you’re writing up your report.)

```{r, echo = FALSE}
q2_data <- df_2020[, c('V201231x','V202119', 'V202120e','V202120i', 'V201024')] %>% 
  mutate(
    Party = ifelse(V201231x == 1 | V201231x == 2 | V201231x == 3, 'Democrat', 'Republican'),
    Ordinal_Difficulty = ifelse(V202119 == 1, 'Not at All', 
                                ifelse(V202119 ==2, 'A Little', 
                                       ifelse(V202119 == 3, 'Moderately', 
                                              ifelse(V202119 == 4, 'Very',
                                                     ifelse(V202119 == 5, 'Extremely', NA)))))
    )%>% 
  filter(
    (V201231x >= 1 & V201231x <= 3) | (V201231x >= 5 & V201231x <= 7),
    V202119 >= 1 & V202119 <= 5
  )

#Rename Columns:
colnames(q2_data) <- c("Political_Affiliation", "Difficulty_Voting", "Difficulty_Getting", "Difficulty_Mailing", "Voting_Method","Party","Ordinal_Difficulty")

#Reorder Columns:
q2_data <- q2_data[,c(1,6,2,7,5,3,4)]


#Establish Order of Ordinal Variables:
q2_data$Ordinal_Difficulty <- factor(q2_data$Ordinal_Difficulty, order = TRUE, 
                                    levels = c("Not at All", "A Little",
                                               "Moderately", "Very",
                                               "Extremely"))


```

```{r}
Dems = sum(q2_data$Party == 'Democrat')
Reps = sum(q2_data$Party == 'Republican')

data1 <- data.frame(
  q2_data %>% 
  group_by(
    Ordinal_Difficulty,
    Party
  ) %>% 
  summarise(
    Democrats = 100*sum(Party =='Democrat')/Dems,
    Republicans = 100*sum(Party == 'Republican')/Reps
  ) %>% 
  mutate(
    Percent = Democrats+Republicans
  )
)

data1 %>% 
  ggplot()+
  aes(x = Ordinal_Difficulty, y= Percent, fill = Party)+
  geom_col(position = 'dodge', alpha = .7)+
  scale_fill_manual(values = c("blue", "red"))+
  labs(
    title    = 'title',
    subtitle = 'subtitle', 
    x        = 'Difficulty Voting',
    y        = 'Percentage of Respondents', 
    fill     = 'Political Affiliation'
  )

tab2 <- q2_data %>%
  group_by(Ordinal_Difficulty) %>%
  summarise(
    Democrat = round((100*length(which(Party == "Democrat"))/Dems), digits = 2),
    Republican = round((100*length(which(Party == "Republican"))/Reps), digits = 2)
  )


knitr::kable(tab2)

```

#Wilcox Test: (non-parametric, ordinal variable, unpaired/ independent groups)

```{r}
wilcox.test(Difficulty_Voting ~ Party, data = q2_data)
```

#T-test: difficulty getting to polling site compared between Democrats and Republicans

#first define a new df that consists of only the values in V202120e that are either 1 or 0 (exclude NA answers)

#distribution of sample means should be approximately normal given sample size even if the distribution of V202120e is binomial 
```{r}
Getting_to_Polls <- q2_data %>%
  filter(
    (Difficulty_Getting >= 0)
  ) %>% 
  select(
    Party,
    Difficulty_Getting
  )

t.test(Getting_to_Polls$Difficulty_Getting ~ Getting_to_Polls$Party)
```


#T-test: difficulty mailing vote compared between Democrats and Republicans

#first define a new df that consists of only the values in V202120i that are either 1 or 0 (exclude NA answers)

#distribution of means should be approximately normal given sample size even if the distribution of V202120i is binomial 
```{r}
Mail<- q2_data %>%
  filter(
    (Difficulty_Mailing >= 0)
  ) %>% 
  select(
    Party,
    Difficulty_Mailing
  )

t.test(Mail$Difficulty_Mailing ~ Mail$Party, conf.level = .95)
```



----------------------------------------------------------------------
3. Are people who believe that science is is important for making government 
decisions about COVID-19 more likely to disapprove of the way their governor 
is handling the pandemic?

```{r}
question3Data <- df_2020[, c("V202310", "V201145", "V201146")]
colnames(question3Data) <- c("ImportanceOfScience", "ApproveGovernorsHandling",
                             "StronglyOrNotStrongly")
question3Data <- question3Data[question3Data$ApproveGovernorsHandling == 1 |
                               question3Data$ApproveGovernorsHandling == 2, ]
question3Data <- question3Data[question3Data$ImportanceOfScience > 0, ]
question3Data$ApproveGovernorsHandling <- question3Data$ApproveGovernorsHandling == 1
question3Data$IsScienceImportant <- question3Data$ImportanceOfScience > 1


tab <- table(question3Data$ApproveGovernorsHandling, question3Data$IsScienceImportant)
row.names(tab) <- c("Disapprove of Governor", "Approve of Governor")
colnames(tab) <- c("Science isn't important", "Science is Important")
tab

difficulty_histogram <- question3Data %>% 
  ggplot() + 
  aes(x = ImportanceOfScience, fill = ApproveGovernorsHandling) + 
  geom_histogram(alpha = .7, position = 'dodge', bins = 5, color = 'black') + 
  scale_fill_manual(values = c("red", "blue"), 
                    labels = c("Disapprove", "Approve"),
                    name = "Approve of Governor's\nHandling of COVID-19") +
  labs(
    title    = 'title',
    subtitle = 'subtitle', 
    x        = 'Importance of Science',
    y        = 'Voters', 
    fill     = 'Approve of Governor'
  )

difficulty_histogram

t.test(question3Data$ApproveGovernorsHandling ~ question3Data$IsScienceImportant, alternative = "g")
```



